---
title: "compas_analysis"
author: "Rami Boughanmi & Maël RIBES"
date: "2023-03-24"
output: html_document:
  number_sections: TRUE
---

# Compas Dataset Analysis

------------------------------------------------------------------------

## Library imports

```{r}

library(CGPfunctions)
library(treemap)
library(ggplot2)
```

## Dataset cleaning

We saw that each row in the original dataset was duplicated for each type of assessment. We have therefore separated the dataset into 3 dataframes corresponding to each of the categories.

```{r}
# Import CSV
compas_all_data <- read.csv("compas-scores-raw.csv")

# Cleaning ethnic codes 
compas_all_data$Ethnic_Code_Text <- sub("African-American", "Afro-Am", compas_all_data$Ethnic_Code_Text)
compas_all_data$Ethnic_Code_Text <- sub("African-Am", "Afro-Am", compas_all_data$Ethnic_Code_Text)
compas_all_data$Ethnic_Code_Text <- sub("Native American", "Native-Am", compas_all_data$Ethnic_Code_Text)

# Calculation of defendants' age and addition of the "age" column to compas_data
compas_all_data$DateOfBirth <- as.POSIXct(paste0(substr(compas_all_data$DateOfBirth,1,6),"19",substr(compas_all_data$DateOfBirth,7,8)), format = "%m/%d/%Y")
compas_all_data$Age <- as.integer(difftime(Sys.time(), compas_all_data$DateOfBirth, units = "days")/365.25)

# Separation of datasets 
RK_recidivism <- subset(compas_all_data, DisplayText == "Risk of Recidivism")
RK_violence <- subset(compas_all_data, DisplayText == "Risk of Violence")
RK_failure_appear <- subset(compas_all_data, DisplayText == "Risk of Failure to Appear")

```

## Risk of violence analysis

As a first step, we chose to study the risk of violence dataset. Let's look at the number of defendants and the distribution of age, sexes, ethnicity etc.

### Ethnicity analysis

```{r}
# Calculation of the number of defendants
nb_defendants <- nrow(RK_violence)

# Extraction of all ethnic groups and their number of individuals
all_ethnicity <- unique(RK_violence$Ethnic_Code_Text)
all_ethnicity_count <- data.frame(ethnicity = character(), count = numeric())
RK_violence_ethnicity <-  subset(RK_violence, select = c(Ethnic_Code_Text, ScoreText)) 
for (ethnicity in all_ethnicity){
  count <- nrow(subset(RK_violence, Ethnic_Code_Text == ethnicity))
  all_ethnicity_count <- rbind(all_ethnicity_count, data.frame(ethnicity = ethnicity, count = count))
}

# Console display of the distribution of different races
print(paste("Total number of defendants :", nb_defendants))
for (ethnicity in all_ethnicity){
  print(paste("distribution of",ethnicity, ":", round((nrow(subset(RK_violence_ethnicity, Ethnic_Code_Text == ethnicity))/nb_defendants)*100,3),"%"))
}

print(all_ethnicity_count)

treemap(all_ethnicity_count, 
        index = "ethnicity", 
        vSize = "count", 
        type = "index",
        title = "Ethnicity count",
        title.legend = "Ethnicity",
        )

PlotXTabs2(
  data = RK_violence_ethnicity,
  y = ScoreText,
  x = Ethnic_Code_Text,
  plottype = "side",
  xlab = "Ethnicity",
  ylab = NULL,
  label.text.size = 2,
  results.subtitle = FALSE,
  sample.size.label = FALSE,
  data.label = "counts",
  legend.title = "Risk of violence",
  legend.position = "right",
  title = "Risk of violence by ethnicity",
  palette = "Set1"
)

PlotXTabs2(
  data = RK_violence_ethnicity,
  y = ScoreText,
  x = Ethnic_Code_Text,
  xlab = "Ethnicity",
  ylab = NULL,
  results.subtitle = FALSE,
  legend.title = "Risk of violence",
  legend.position = "right",
  title = "Risk of violence by ethnicity",
  palette = "Set1"
)

```

The majority of defendants belong to the ethnic groups of African Americans (44.406%) and Caucasians (35.802%).
The other ethnic groups represent a much smaller proportion.
The percentage difference between races may indicate possible discrimination or bias in the system and may also affect our future analyses.

### Ages analysis

```{r}

all_ages = unique(RK_violence$Age)

# Console display of the distribution of different ages
for (age in all_ages){
  print(paste("Distribution of people who are ", age, "years old :", round((nrow(subset(RK_violence, Age == age))/nb_defendants)*100, 3),"%"))
}

# Create a vector with the number of defendants per age
def_by_age <- sapply(all_ages, function(age) nrow(subset(RK_violence, Age == age)))

# Display a bar chart for the distribution of ages
barplot(def_by_age, names.arg = all_ages, xlab = "Age", ylab = "Number of defendants", main = "Distribution of defendants' ages")

# Create a data frame with the distribution of ages
age_distribution <- data.frame(age = all_ages, percentage = sapply(all_ages, function(x) {nrow(subset(RK_violence, Age == x))/nb_defendants}))

# Create a histogram of age distribution
ggplot(age_distribution, aes(x=age, y=percentage)) + 
  geom_bar(stat="identity") +
  labs(title="Age distribution of the defendants", x="Age", y="Pourcentage") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

```

We observe that there is still a fairly uniform distribution among the different age groups,
#however, we still have a concentration of defendants in the age group of 30-40 years.

### Sexes analysis

```{r}

all_sexes = unique(RK_violence$Sex_Code_Text)

# Console display of the distribution of different sexes
for (sex in all_sexes){
  print(paste("Distribution of",sex,":", round((nrow(subset(RK_violence, Sex_Code_Text == sex))/nb_defendants)*100, 3),"%"))
}

# Create a data frame with the sex distribution
sex_distribution <- data.frame(sex = all_sexes, percentage = sapply(all_sexes, function(x) {nrow(subset(RK_violence, Sex_Code_Text == x))/nb_defendants}))

# Pie chart of sex distribution
ggplot(sex_distribution, aes(x = "", y = percentage, fill = sex)) + 
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Sex distribution in the dataset")+ 
  theme_void() +
  geom_text(aes(label = paste0(round(percentage*100), "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette="Set1")
```

We observe here that the majority sex of individuals in the compas dataset are men (78%).
